Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-19 paper/spectraldensity.tex
review pass through beginning of section 3

To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we first approximate the function 
$s\mapsto g_{\sigma}(t- s)$ by a truncated
Chebyshev expansion:
%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation. For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation. In every $t \in \mathbb{R}$ we expand 
\begin{equation}
    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
    \label{equ:matrix-expansion}
\end{equation}
with the expansion coefficients $\mu_0(t), \dots, \mu_m(t) \in \mathbb R$ and the Chebyshev polynomials of the first kind
$T_1,\ldots,T_m$. Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, one has that
$T_l(s) = \cos(l \cdot \arccos(s))$ satisfies the 
three-term recurrence relation
\begin{equation}
    T_0(s) = 1,\quad  T_1(s) = s,\quad 
    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
    \label{equ:chebyshev-recurrence}
\end{equation}
Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
\begin{equation}
    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
    \label{equ:expanded-spectral-density}
\end{equation}

For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum \cite{zhou-2011-bounding-spectrum}. The usual affine transform
$\tau(t) = \frac{2t - a - b}{b - a}$
then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} From now on, we will assume without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.

@@

@@

@@

Similar observations for the case when $n_{\mtx{\Psi}}$ is nonzero lead to the following algorithm.

@@

\begin{proof}
    For any numbers $a, b \in \mathbb{R}$ it holds
    \begin{equation}
    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
    \end{equation}
    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
    we have
    \begin{equation}
        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
    \end{equation}
    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ we can apply the result from \refthm{thm:chebyshev-error} to get
    \begin{align}
        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
    \end{align}
    from which follows the result with some minor simplifications.

>>>>

To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we approximate the function 
$s\mapsto g_{\sigma}(t- s)$ by a truncated
Chebyshev expansion $g_{\sigma}^{(m)}(t-s) = 
\sum_{l=0}^{m} \mu_l(t) T_l(s)$
with the expansion coefficients $\mu_l(t) \in \mathbb R$ and the Chebyshev polynomials
$T_l(s) = \cos(l \cdot \arccos(s))$ for $l = 0,1,\ldots,m$.
Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation. For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation. In every $t \in \mathbb{R}$ we expand 
\begin{equation}
    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
    \label{equ:matrix-expansion}
\end{equation}
\textcolor{red}{??? MOVE BELOW! This is not at a good place
For later purposes, we recall the well-known
three-term recurrence relation
\begin{equation}
    T_0(s) = 1,\quad  T_1(s) = s,\quad 
    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
    \label{equ:chebyshev-recurrence}
\end{equation}}
Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
\begin{equation}
    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
    \label{equ:expanded-spectral-density}
\end{equation}
\begin{remark}
For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum of $A$~\cite{zhou-2011-bounding-spectrum}. The usual affine transform
$\tau(t) = \frac{2t - a - b}{b - a}$
then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on, without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
\end{remark}

@@

\subsubsection{Preservation of non-negativity}

@@

\subsection{Computation of expansion coefficients}

@@

Similar observations for the case when $n_{\mtx{\Psi}}$ is nonzero lead to the following algorithm.
\textcolor{red}{??? Lines 1 and 2 are quite confusing. Say more explicitly how $\mu$ is computed. Is Algorithm 3.3 applied twice? ???}

@@

\begin{proof}
    For any numbers $a, b \in \mathbb{R}$ it holds
    \begin{equation}
    | a^2 - b^2 | = | (a + b)(a - b) | = | a + b | | a - b | \leq (2 | a | + | a - b |)  | a - b |
    \end{equation}
    Therefore, omitting the arguments and using $g_{\sigma} = (\sqrt{g_{\sigma}})^2$ and $\underline{g}_{\sigma}^{(m)} = (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2$.
    we have
    \begin{equation}
        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
    \end{equation}
    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$ and $\sqrt{g_{\sigma}} \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ \textcolor{red}{we can apply the result from \refthm{thm:chebyshev-error} ??? the reference seems to be incorrect ???} to get
    \begin{align}
        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}},
    \end{align}
    from which follows the result with some minor simplifications.
